Estimation of the perceived time of presence of sources in urban acoustic environment using deep learning techniques

### 2019-06-04 paper/paperV1.tex
JFP proofread

\section{Introduction}
\label{sec:intro}
In urban areas, the decrease of quality of sound environments resulting from urbanization is a major cause of annoyance, and has been identified as an important health factor. Consequently, the 2002/49/CE European directive~\cite{ec2002} requires the development of noise maps in large cities. These maps aim at providing useful tools for urban planning and noise reduction purposes. As an alternative to the noise-only paradigm, the soundscape approach defined in the ISO 12913-1~\cite{iso2014} norm as "the acoustic environment as perceived and understood and/or experienced by people and/or society, in context" extends the characterization of sound environments by considering the possible positive influence of some of its components.\\

The assessment of soundscape quality through standardized perceptual descriptors has been extensively studied~\cite{viollon2000, axelsson2010, cain2013, jeon2018, aletta2016}. Aletta et al.~\cite{aletta2016} did a review of soundscape descriptors and conclude that the potential soundscape descriptors identified so far seem to converge towards a two-dimensional soundscape model of perceived affective quality where one dimension corresponds to the pleasantness of the soundscape, and the orthogonal one corresponds to its enventfulness~\cite{axelsson2010, aumond2017, delaitre2014}. In this paper, we will focus on the first affective dimension which is pleasantness.\\

Pleasantness is mainly linked to the perceived loudness~\cite{blauert1997, jekosch2004} but it also depends on the activity of sources of interest~\cite{nilsson2007, perez2012}. While the exact taxonomy of sources used differs across studies~\cite{guastavino2007, gygi2007, salamon2014, brown2011}, three major source types are found to contribute to pleasantness in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010} where traffic noise, human voices and birds calls can respectively be used as a proxy of these three types~\cite{lavandier2006, ricciardi2014, aumond2017}. In existing models, source activity can be quantified by the dominance~\cite{hong2017}, the emergence~\cite{guastavino2006} or the time of presence~\cite{ricciardi2014}. Though, the perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the production of perceptually motivated noise maps.\\

@@

In this paper, we do so by developing new indicators relying on source recognition models based on deep learning techniques, which is a growing interest in the machine listening community~\cite{cakir2015, salamon2017-2}. Large amounts of data with task-specific annotations are needed for these architectures to learn to extract relevant information from recordings. To the best of our knowledge such databases do not currently exist in the literature in the case of source recognition with perceptual activity labels. We propose in this paper to study the potential of the use of simulated corpora~\cite{rossignol2015, salamon2017} for such purpose.\\

@@

The 45~s sound scenes are then replicated using \textit{simScene}~\cite{rossignol2015}. \textit{simScene} is a software library in Matlab that allows the simulation of sound scenes as the additive composition of sound sources, using a database of recordings associated to these sources. Here the database is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources. The original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are chosen for perceptual comparison, as they explore diverse real-life situations with respect to environment categorization. All 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL.\\

@@

Adjustments are made on the variance of the considered properties to better cover possible scenarios. Voice events consist mostly of read English recordings, thus their EBR is reduced to improve realism. An additional \textit{square} ambiance is added with properties derived empirically from available data. Diverse new scenarios are then generated by sampling the background and event properties distributions, resulting in 40 scenes per ambiance for a total of 200 scenes. For each scene the perceptual time of presence for the traffic, voice and bird sources is estimated using the indicators proposed in~\cite{gontier2018}. Ideally, sound scenes should cover the resulting 3-dimensional space in a homogeneous way. Though, the time of presence is not a parameter of the simulation process and thus the distribution of the scene population in this space is not controlled. As an alternative, 75 scenes are selected that maximize the minimum pairwise distance in the 3-dimensional space. This is shown in Figure~\ref{fig:tvb_pres}. The playback sound level is also conditioned to the ambiance according to typical sound levels in urban context, and ranges from 46.6~dB SPL to 77.1~dB SPL.\\

@@

These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}. While this work focuses on the application of pleasantness prediction, including other dimensions can be useful to verify the relevance of simulated data in relation to previous work on real scenarios.\\

The perception of source activity can be linked to the notions of dominance, time of presence, or emergence. We choose the perceptual time of presence, that is the percentage of time for which the corresponding source is heard, as the activity descriptor for this study. The influence of traffic on pleasantness is further separated into that of background and salient events. Thus, 4 additional questions are presented to the participants and evaluated on the same 11 point scales:
\begin{itemize}
\item Sound level of passing vehicles: \textit{Very low - Very high} (\textit{Tr\`es faible - Tr\`es fort}) - $L_{T, p}$,
\item Time of presence of Traffic, Voices, Birds: \textit{Never - Continuously} (\textit{Jamais - Continuellement}), resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
where $p$ denotes a perceptual evaluation.\\

Prior to the test, a short verbal introduction is given to the participants and the interface is introduced to ensure that the quantities are well understood. Although the corpus is comprised of 100 sound scenes, participants only evaluate 50 scenes: all listen to the 6 recorded and 19 replicated sound scenes, then to 25 of the 75 simulated with new scenarios according to a balanced incomplete block design~\cite{dagnelie2003}. The selection of simulated scenes is done so that all scenes in the sub-corpus are evaluated by the same number of participants. All participants are first presented with the most quiet then most noisy of the recorded scenes (resp. P3 and P15). A random listening order is generated for each subject to control ordering effects for the remaining of the test. Participants can listen to each scene once, and have to answer all questions and listen to the full extract before proceeding to the next.\\

The scenes are played at a given sound level as discussed in Section~\ref{sec:data_corp}, through the same computer and sound card configurations. Beyerdynamics DT-990 Pro headphones are used for all participants. These headphones are calibrated in a semi-anechoic chamber where a relation between the playback sound level and the sound card output voltage is obtained for pink noise. From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by listeners.\\

@@

The perceptual responses of the 6 recorded and corresponding replicated scenes are first compared. Table~\ref{tab:ogrep} shows the mean differences between assessments for pairs of scenes with equivalent scenarios. Wilcoxon signed-rank tests~\cite{wilcoxon1945} are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold. As the data is discretely distributed, zero differences between paired samples are included using Pratt's modification of the test~\cite{pratt1959}. On the general scales (P, L, OL, I and C), all mean differences are lower than 2 points on the 11-point Likert scale (catherine?). Though, significant differences are outlined that can be linked to corresponding discrepancies in source-specific parameters. The highest difference (-5.22) is found for the assessment of the time of presence of traffic in the location P4. For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half. Replicating this scene using simScene imposes a constant sound level for background sources. Thus, the background traffic is louder in the replicated scene than it is in the recording for about half of its duration. To a lesser extent the same issue explains the large difference (-4) in the assessed time of presence of voices in the location P3. Discrepancies on source-specific scales can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy. For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness. Overall, no consistent difference between the perception of recorded and replicated scenes emerges for the studied points. Thus, the proposed hypothesis that recorded and replicated scenes generate the same perceptual responses cannot be rejected.\\

@@

To validate the use of simulated sound scenes with new scenarios as well as reduced source complexity in the development of prediction models, the perceptual space generated by the experiment's five general scales (pleasantness, liveliness, overall loudness, interest and calmness) is studied. It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses averaged along participants. No standardization is applied to the data. Figure~\ref{fig:pspace_rec} and Figure~\ref{fig:pspace_sim} compare the results for scenes based on recordings (n=25) and new scenarios (n=75) respectively. The resulting spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two subcorpora. For both sets the variance explained by the first two components is similar, resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%. Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson2010, cain2013}. Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities. Additionally, the assessments averaged on all subjects are projected onto the PCA space for each scene and represented in Figure~\ref{fig:pspace_p1} as red dots for simulated scenes corpus and blue dots for recorded and replicated scenes. The space covered by scenes based on new scenarios covers that of the studied real-life environments, further demonstrating the effectiveness of the scene generation procedure in terms of diversity of scenarios.\\

Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace_rec} using arrows. The standard deviation of assessments is represented by ellipses for one location (P1). The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes, available on the companion website~\footnote{URL}. This illustrates the results in Table~\ref{tab:ogrep} where perceptual assessments were not found to differ significantly overall.\\

@@

Based on previous studies~\cite{aumond2017, gontier2018, ricciardi2014}, several indicators are identified that correlate well with perceptual parameters. These indicators can be used to develop baseline physical models of pleasantness in the design of a deep learning architecture for sound source recognition.\\

@@

where $r$ is the Pearson's correlation coefficient, $s$ denotes the sound source, $N_s$ is the number of sources in the taxonomy and equals 3 in this study, $t$ is the time frame and $T_{s, p}$ corresponds to the perceptual time of presence assessments averaged per scene. This indicator evaluates the presence or absence of a given source in a small time frame. First, frequency bands for which the source is emergent by more than the $\alpha$ threshold value are isolated. Then, the source is considered present if the emergence for these bands is on average greater than the $\beta$ threshold. A time of presence estimation is obtained for each source by averaging along the $N_t$ time frames composing the scene. The optimal threshold values are optimized once in this study on the whole perceptual experiment corpus, and found as $\alpha_{opt} = -14dB$ and $\beta_{opt} = -7dB$. This means that to be considered heard in a given time frame, the source has to be emergent by more than $-7dB$ on average on third-octave bands where its level is at least $14dB$ below other sources.\\

@@

The generation procedure is the same for both datasets, and is equivalent to that of the 75 scenes in the simulated sub-corpus considered for the perceptual test in Section~\ref{sec:data_corp}. Only traffic, human voices and birds sources compose the sound scenes. To generate the scenes, it is important to ensure a complete independance of the training and test set. For that, the isolated samples database is split in the same proportions as the two datasets: two-thirds for the development set and the remaining one-third for the evaluation set. As a result, while the generation process is the same for both corpora, the scenarios and audio samples are different.\\

The developped model should extract relevant source information from a representation of the audio signal. Typically, spectral representations such as mel bands are preferred to the raw audio waveform because of the regularities they underline in the signal. Here, the third-octave spectrum is considered as it is commonly used in acoustic monitoring applications~\cite{ardouin2018, gontier2017}. Third-octave spectra are computed for 125~ms frames and 29 frequency bands in the $20~Hz - 12.5~kHz$ range. Inputs are then obtained by splitting the resulting spectrograms into texture windows of 1~s duration, which is considered a relevant scale for perception. The spectral blocks of dimension 29x8 are then processed independently in this study.\\

Target outputs are associated to each input frame to train the model. In~\cite{gontier2018} the time of presence of traffic, voices and birds was successfully approximated by the $\hat T_s(\alpha, \beta)$ indicator. Thus, the source presence values given by this indicator can be used as weak perceptual time of presence labels. Alternatively, it is possible to perform sound level estimation using ground truth labels provided by the simulation process before computing $\hat T_s(\alpha_{opt}, \beta_{opt})$. Though, training a deep learning model for sound level regression is more difficult: it requires the architecture to perform implicit source separation, whereas presence detection does not need this information to be extracted from the data. This will not be discussed here as it did not result in better time of presence estimations in our experiments.\\

A deep learning model is implemented and trained for the source presence prediction task, with the architecture shown in Figure~\ref{fig:deep_arch}. The model includes 4 blocks of convolutional layers followed by leaky rectified linear unit (LeakyReLU) activations of expression $y = max(0.1x, x)$. The convolutional layers have respectively 128, 64, 32 and 32 output maps, \textit{i.e.} channels, and a common kernel size of 5x5. The output of the last block is flattened then goes through a fully connected layer with output size 3. A final sigmoid activation is used to obtain outputs in the 0-1 range. The resulting 3 values correspond to the presence of traffic, voices and birds in the 1~s frame respectively. During training these values are directly compared to presence labels given by $\hat T_s(\alpha_{opt}, \beta_{opt})$ using a binary cross-entropy cost function:
\begin{equation}
BCE(y, \hat y) = -\sum_s y_s log\left(\hat y_s\right) + (1-y_s) log\left(1-\hat y_s\right)
\end{equation}
where $s$ is the source, $y_s$ and $\hat y_s$ are the target and predicted presence for source $s$ in the 0-1 range. This loss function is minimized using the Adam algorithm~\cite{kingma2015} on batches of 1~s examples. During evaluation, a threshold of 0.5 is independently applied to the 3 outputs to obtain a binary presence value for each source: each source is considered absent when the model outputs a value lower than 0.5 and present when it outputs 0.5 or higher. The time of presence estimation is then obtained by averaging presence labels of all 1~s time frames corresponding to the same scene.\\

@@

We first study the relations between perceptual scales with respect to existing work. Table~\ref{tab:percc} shows the Pearson's correlation coefficients between pairs of parameters with assessments averaged for each scene (n=100). The resulting values are consistent with the literature~\cite{aumond2017, gontier2018}, both between general scales as discussed in Section~\ref{sec:data_val} and for the relations to source contributions. Indeed, pleasantness (P) is mainly influenced negatively by overall loudness (OL) and traffic and positively by birds. In previous studies a small positive contribution of voices to pleasantness was found, while no direct relation is visible from the data gathered. This can be explained by the choice of speech samples used during generation, which consist of read audiobooks extracts and thus may sound unnatural in the considered urban environments. Liveliness (L) is here negatively though weakly (r=-0.36) correlated with the presence of birds, while no significant relation was found in previous studies. However, its correlation with the presence of human voices is the strongest (r=0.60), as suggested by~\cite{axelsson2010}.\\

Relations between source-specific parameters are also weak with the exception of traffic being negatively correlated with birds (r=-0.42). The perceived sound level of passing vehicles $L_{T, p}$ is a potential alternative to the time of presence of traffic $T_{T, p}$, as it displays higher correlations with general scales and lower correlations with other source-specific parameters.\\

@@

Because of the collinearity between $L_{T, p}$ and $T_{T, p}$, both models have very similar formal expressions, with important negative contributions of the overall loudness and traffic activity. A small positive contribution of the time of presence of voices is also found despite no significant correlation underlined in Table~\ref{tab:percc}.\\

@@

The analysis of physical indicators presented in Section~\ref{sec:methods_inds} is done using the arithmetic mean of subjective assessments over all participants for replicated and simulated scenes. The 6 recorded scenes are excluded from the study as the ground truth source composition is unknown. Two scenes containing only one source are also excluded for implementation concerns, resulting in $n=92$ scenes.\\

Table~\ref{tab:physc} shows the Pearson's correlation coefficients between computed physical indicators and perceptual assessments. First, all global sound level indicators correlate well (r>0.9) with the perceived overall loudness. Regarding source-specific perceptual parameters, the $L_{eq, s}$ correlates consistently well with the $T_{s, p}$ of corresponding source (r=0.71). Conversely, the emergence $\Delta L_s$ fails to represent the perceived bird activity, and correlations are weak for other sources. The proposed estimates of the time of presence $\hat T_s(\alpha_{opt}, \beta_{opt})$ show strong correlations to their perceptual counterparts (r>0.8), though this is expected as $\alpha_{opt}$ and $\beta_{opt}$ are optimized to this aim. They also display good source discrimination properties, as no significant correlation is found between voices and birds, and perceptual assessments of traffic were correlated with those of other sources in Table~\ref{tab:percc}. They are also better predictors than the $TFSD_{500Hz, 1s}$ and $TFSD_{4kHz, 1/8s}$ indicators for voice and bird activity in this corpus. Interestingly, the perceived sound level of passing vehicles $L_{T, p}$ is best represented by the $L_{10}$. In perceptual models including the sound level of traffic events, the global $L_{10}$ may thus be a sufficient predictor without a source separation and level estimation algorithm.\\

@@

Subsequently, multilinear models of pleasantness are constructed based on the results of the analysis of correlations. As in Section~\ref{sec:results_perc}, all combinations of the proposed physical variables are considered, and a VIF check is performed on predictors to ensure that no multi-collinearity between predictors exists in a model. On the present corpus, the best model in terms of $R^2_{adj}$ is:
\begin{equation}
\hat P_{1, \varphi} = 16.71 - 0.18~L_{50} + 1.01~\hat T_B(\alpha_{opt}, \beta_{opt}) \label{eq:pphi}
\end{equation}
where $\varphi$ indicates a model from physical variables. Indicators related to traffic and voices activity are absent compared to the perceptual models in Section~\ref{sec:results_perc}. This is because the $L_{50}$ used as a predictor is correlated to both traffic parameters $L_{T, p}$ and $T_{T, p}$ that appeared in perceptual models $\hat P_{1, p}$ (eq.\ref{eq:pp1}) and $\hat P_{2, p}$ (eq.\ref{eq:pp2}) respectively, and no strong contribution of objective variables representative of voices to pleasantness is identified in this corpus.\\

@@

Pleasantness prediction for simulated scenes using the detection model is as effective as the best baseline model from Section~\ref{sec:results_phys}, and outperforms previous models relying on physical indicators directly computed on the audio signal. Though, perceptual baseline models presented in Section~\ref{sec:results_perc} still display better overall performance metrics for all considered sub-corpora.\\

>>>>

\section{Introduction}
\label{sec:intro}
In urban areas, the decrease of quality of sound environments resulting from urbanization is a major cause of annoyance, and has been identified as an important health factor. Consequently, the 2002/49/CE European directive~\cite{ec2002} requires the development of noise maps in large cities. These maps aim at providing useful tools for urban planning and noise reduction purposes. As an alternative to the noise-only paradigm, the soundscape approach defined in the ISO 12913-1~\cite{iso2014} norm as "the acoustic environment as perceived and understood and/or experienced by people and/or society, in context" extends the characterization of sound environments by considering the possible positive influence of some of its components.

The assessment of soundscape quality through standardized perceptual descriptors has been extensively studied~\cite{viollon2000, axelsson2010, cain2013, jeon2018, aletta2016}. Aletta et al.~\cite{aletta2016} did a review of soundscape descriptors and conclude that the potential soundscape descriptors identified so far seem to converge towards a two-dimensional soundscape model of perceived affective quality where one dimension corresponds to the pleasantness of the soundscape, and the orthogonal one corresponds to its enventfulness~\cite{axelsson2010, aumond2017, delaitre2014}. In this paper, we will focus on the first affective dimension which is pleasantness. Pleasantness is mainly linked to the perceived loudness~\cite{blauert1997, jekosch2004} but it also depends on the activity of sources of interest~\cite{nilsson2007, perez2012}. While the exact taxonomy of sources used differs across studies~\cite{guastavino2007, gygi2007, salamon2014, brown2011}, three major source types are found to contribute to pleasantness in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010} where traffic noise, human voices and birds calls can respectively be used as a proxy of these three types~\cite{lavandier2006, ricciardi2014, aumond2017}. In existing models, source activity can be quantified by the dominance~\cite{hong2017}, the emergence~\cite{guastavino2006} or the time of presence~\cite{ricciardi2014}. Though, the perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the production of perceptually motivated noise maps.

@@

In this paper, we do so by developing new indicators relying on source recognition models based on deep learning techniques, which is a growing interest in the machine listening community~\cite{cakir2015, salamon2017-2}. Large amounts of data with task-specific annotations are needed for these architectures to learn to extract relevant information from recordings. To the best of our knowledge such databases do not currently exist in the literature in the case of source recognition with perceptual activity labels. We propose in this paper to study the potential of the use of simulated corpora~\cite{rossignol2015, salamon2017} for such purpose.

@@

The 45~s sound scenes are then replicated using \textit{simScene}~\cite{rossignol2015}. \textit{simScene} is a software library in Matlab that allows the simulation of sound scenes as the additive composition of sound sources, using a database of recordings associated to these sources. Here the database is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources. The original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are chosen for perceptual comparison, as they explore diverse real-life situations with respect to environment categorization. All 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL.

@@

Adjustments are made on the variance of the considered properties to better cover possible scenarios. Voice events consist mostly of read English recordings, thus their EBR is reduced to improve realism. An additional \textit{square} ambiance is added with properties derived empirically from available data. Diverse new scenarios are then generated by sampling the background and event properties distributions, resulting in 40 scenes per ambiance for a total of 200 scenes. For each scene the perceptual time of presence for the traffic, voice and bird sources is estimated using the indicators proposed in~\cite{gontier2018}. Ideally, sound scenes should cover the resulting 3-dimensional space in a homogeneous way. Though, the time of presence is not a parameter of the simulation process and thus the distribution of the scene population in this space is not controlled. As an alternative, 75 scenes are selected that maximize the minimum pairwise distance in the 3-dimensional space. This is shown in Figure~\ref{fig:tvb_pres}. The playback sound level is also conditioned to the ambiance according to typical sound levels in urban context, and ranges from 46.6~dB SPL to 77.1~dB SPL.

@@

These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}. While this work focuses on the application of pleasantness prediction, including other dimensions can be useful to verify the relevance of simulated data in relation to previous work on real scenarios.

The perception of source activity can be linked to the notions of dominance, time of presence, or emergence. We choose the perceptual time of presence, that is the percentage of time for which the corresponding source is heard, as the activity descriptor for this study. The influence of traffic on pleasantness is further separated into that of background and salient events. Thus, 4 additional questions are presented to the participants and evaluated on the same 11 point scales:
\begin{itemize}
\item Sound level of passing vehicles: \textit{Very low - Very high} (\textit{Tr\`es faible - Tr\`es fort}) - $L_{T, p}$,
\item Time of presence of Traffic, Voices, Birds: \textit{Never - Continuously} (\textit{Jamais - Continuellement}), resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
where $p$ denotes a perceptual evaluation.

Prior to the test, a short verbal introduction is given to the participants and the interface is introduced to ensure that the quantities are well understood. Although the corpus is comprised of 100 sound scenes, participants only evaluate 50 scenes: all listen to the 6 recorded and 19 replicated sound scenes, then to 25 of the 75 simulated with new scenarios according to a balanced incomplete block design~\cite{dagnelie2003}. The selection of simulated scenes is done so that all scenes in the sub-corpus are evaluated by the same number of participants. All participants are first presented with the most quiet then most noisy of the recorded scenes (resp. P3 and P15). A random listening order is generated for each subject to control ordering effects for the remaining of the test. Participants can listen to each scene once, and have to answer all questions and listen to the full extract before proceeding to the next.

The scenes are played at a given sound level as discussed in Section~\ref{sec:data_corp}, through the same computer and sound card configurations. Beyerdynamics DT-990 Pro headphones are used for all participants. These headphones are calibrated in a semi-anechoic chamber where a relation between the playback sound level and the sound card output voltage is obtained for pink noise. From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by listeners.

@@

The perceptual responses of the 6 recorded and corresponding replicated scenes are first compared. Table~\ref{tab:ogrep} shows the mean differences between assessments for pairs of scenes with equivalent scenarios. Wilcoxon signed-rank tests~\cite{wilcoxon1945} are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold. As the data is discretely distributed, zero differences between paired samples are included using Pratt's modification of the test~\cite{pratt1959}. On the general scales (P, L, OL, I and C), all mean differences are lower than 2 points on the 11-point Likert scale. Though, significant differences are outlined that can be linked to corresponding discrepancies in source-specific parameters. The highest difference (-5.22) is found for the assessment of the time of presence of traffic in the location P4. For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half. Replicating this scene using simScene imposes a constant sound level for background sources. Thus, the background traffic is louder in the replicated scene than it is in the recording for about half of its duration. To a lesser extent the same issue explains the large difference (-4) in the assessed time of presence of voices in the location P3. Discrepancies on source-specific scales can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy. For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness. Overall, no consistent difference between the perception of recorded and replicated scenes emerges for the studied points. Thus, the proposed hypothesis that recorded and replicated scenes generate the same perceptual responses cannot be rejected.

@@

To validate the use of simulated sound scenes with new scenarios as well as reduced source complexity in the development of prediction models, the perceptual space generated by the experiment's five general scales (pleasantness, liveliness, overall loudness, interest and calmness) is studied. It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses averaged along participants. No standardization is applied to the data. Figure~\ref{fig:pspace_rec} and Figure~\ref{fig:pspace_sim} compare the results for scenes based on recordings (n=25) and new scenarios (n=75) respectively. The resulting spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two subcorpora. For both sets the variance explained by the first two components is similar, resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%. Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson2010, cain2013}. Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities. Additionally, the assessments averaged on all subjects are projected onto the PCA space for each scene and represented in Figure~\ref{fig:pspace_p1} as red dots for simulated scenes corpus and blue dots for recorded and replicated scenes. The space covered by scenes based on new scenarios covers that of the studied real-life environments, further demonstrating the effectiveness of the scene generation procedure in terms of diversity of scenarios.

Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace_rec} using arrows. The standard deviation of assessments is represented by ellipses for one location (P1). The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes, available on the companion website~\footnote{URL}. This illustrates the results in Table~\ref{tab:ogrep} where perceptual assessments were not found to differ significantly overall.

@@

Based on previous studies~\cite{aumond2017, gontier2018, ricciardi2014}, several indicators are identified that correlate well with perceptual parameters. These indicators can be used to develop baseline physical models of pleasantness in the design of a deep learning architecture for sound source recognition.

@@

where $r$ is the Pearson's correlation coefficient, $s$ denotes the sound source, $N_s$ is the number of sources in the taxonomy and equals 3 in this study, $t$ is the time frame and $T_{s, p}$ corresponds to the perceptual time of presence assessments averaged per scene. This indicator evaluates the presence or absence of a given source in a small time frame. First, frequency bands for which the source is emergent by more than the $\alpha$ threshold value are isolated. Then, the source is considered present if the emergence for these bands is on average greater than the $\beta$ threshold. A time of presence estimation is obtained for each source by averaging along the $N_t$ time frames composing the scene. The optimal threshold values are optimized once in this study on the whole perceptual experiment corpus, and found as $\alpha_{opt} = -14dB$ and $\beta_{opt} = -7dB$. This means that to be considered heard in a given time frame, the source has to be emergent by more than $-7dB$ on average on third-octave bands where its level is at least $14dB$ below other sources.

@@

The generation procedure is the same for both datasets, and is equivalent to that of the 75 scenes in the simulated sub-corpus considered for the perceptual test in Section~\ref{sec:data_corp}. Only traffic, human voices and birds sources compose the sound scenes. To generate the scenes, it is important to ensure a complete independance of the training and test set. For that, the isolated samples database is split in the same proportions as the two datasets: two-thirds for the development set and the remaining one-third for the evaluation set. As a result, while the generation process is the same for both corpora, the scenarios and audio samples are different.

The developped model should extract relevant source information from a representation of the audio signal. Typically, spectral representations such as mel bands are preferred to the raw audio waveform because of the regularities they underline in the signal. Here, the third-octave spectrum is considered as it is commonly used in acoustic monitoring applications~\cite{ardouin2018, gontier2017}. Third-octave spectra are computed for 125~ms frames and 29 frequency bands in the $20~Hz - 12.5~kHz$ range. Inputs are then obtained by splitting the resulting spectrograms into texture windows of 1~s duration, which is considered a relevant scale for perception. The spectral blocks of dimension 29x8 are then processed independently in this study.

Target outputs are associated to each input frame to train the model. In~\cite{gontier2018} the time of presence of traffic, voices and birds was successfully approximated by the $\hat T_s(\alpha, \beta)$ indicator. Thus, the source presence values given by this indicator can be used as weak perceptual time of presence labels. Alternatively, it is possible to perform sound level estimation using ground truth labels provided by the simulation process before computing $\hat T_s(\alpha_{opt}, \beta_{opt})$. Though, training a deep learning model for sound level regression is more difficult: it requires the architecture to perform implicit source separation, whereas presence detection does not need this information to be extracted from the data. This will not be discussed here as it did not result in better time of presence estimations in our experiments.

A deep learning model is implemented and trained for the source presence prediction task, with the architecture shown in Figure~\ref{fig:deep_arch}. The model includes 4 blocks of convolutional layers followed by leaky rectified linear unit (LeakyReLU) activations of expression $y = max(0.1x, x)$. The convolutional layers have respectively 128, 64, 32 and 32 output maps, \textit{i.e.} channels, and a common kernel size of 5x5. The output of the last block is flattened then goes through a fully connected layer with output size 3. A final sigmoid activation is used to obtain outputs in the 0-1 range. The resulting 3 values correspond to the presence of traffic, voices and birds in the 1~s frame respectively. During training these values are directly compared to presence labels given by $\hat T_s(\alpha_{opt}, \beta_{opt})$ using a binary cross-entropy cost function:
\begin{equation}
BCE(y, \hat y) = -\sum_s y_s log\left(\hat y_s\right) + (1-y_s) log\left(1-\hat y_s\right)
\end{equation}
where $s$ is the source, $y_s$ and $\hat y_s$ are the target and predicted presence for source $s$ in the 0-1 range. This loss function is minimized using the Adam algorithm~\cite{kingma2015} on batches of 1~s examples. During evaluation, a threshold of 0.5 is independently applied to the 3 outputs to obtain a binary presence value for each source: each source is considered absent when the model outputs a value lower than 0.5 and present when it outputs 0.5 or higher. The time of presence estimation is then obtained by averaging presence labels of all 1~s time frames corresponding to the same scene.

@@

We first study the relations between perceptual scales with respect to existing work. Table~\ref{tab:percc} shows the Pearson's correlation coefficients between pairs of parameters with assessments averaged for each scene (n=100). The resulting values are consistent with the literature~\cite{aumond2017, gontier2018}, both between general scales as discussed in Section~\ref{sec:data_val} and for the relations to source contributions. Indeed, pleasantness (P) is mainly influenced negatively by overall loudness (OL) and traffic and positively by birds. In previous studies a small positive contribution of voices to pleasantness was found, while no direct relation is visible from the data gathered. This can be explained by the choice of speech samples used during generation, which consist of read audiobooks extracts and thus may sound unnatural in the considered urban environments. Liveliness (L) is here negatively though weakly (r=-0.36) correlated with the presence of birds, while no significant relation was found in previous studies. However, its correlation with the presence of human voices is the strongest (r=0.60), as suggested by~\cite{axelsson2010}.

Relations between source-specific parameters are also weak with the exception of traffic being negatively correlated with birds (r=-0.42). The perceived sound level of passing vehicles $L_{T, p}$ is a potential alternative to the time of presence of traffic $T_{T, p}$, as it displays higher correlations with general scales and lower correlations with other source-specific parameters.

@@

Because of the collinearity between $L_{T, p}$ and $T_{T, p}$, both models have very similar formal expressions, with important negative contributions of the overall loudness and traffic activity. A small positive contribution of the time of presence of voices is also found despite no significant correlation underlined in Table~\ref{tab:percc}.

@@

The analysis of physical indicators presented in Section~\ref{sec:methods_inds} is done using the arithmetic mean of subjective assessments over all participants for replicated and simulated scenes. The 6 recorded scenes are excluded from the study as the ground truth source composition is unknown. Two scenes containing only one source are also excluded for implementation concerns, resulting in $n=92$ scenes.

Table~\ref{tab:physc} shows the Pearson's correlation coefficients between computed physical indicators and perceptual assessments. First, all global sound level indicators correlate well (r>0.9) with the perceived overall loudness. Regarding source-specific perceptual parameters, the $L_{eq, s}$ correlates consistently well with the $T_{s, p}$ of corresponding source (r=0.71). Conversely, the emergence $\Delta L_s$ fails to represent the perceived bird activity, and correlations are weak for other sources. The proposed estimates of the time of presence $\hat T_s(\alpha_{opt}, \beta_{opt})$ show strong correlations to their perceptual counterparts (r>0.8), though this is expected as $\alpha_{opt}$ and $\beta_{opt}$ are optimized to this aim. They also display good source discrimination properties, as no significant correlation is found between voices and birds, and perceptual assessments of traffic were correlated with those of other sources in Table~\ref{tab:percc}. They are also better predictors than the $TFSD_{500Hz, 1s}$ and $TFSD_{4kHz, 1/8s}$ indicators for voice and bird activity in this corpus. Interestingly, the perceived sound level of passing vehicles $L_{T, p}$ is best represented by the $L_{10}$. In perceptual models including the sound level of traffic events, the global $L_{10}$ may thus be a sufficient predictor without a source separation and level estimation algorithm.

@@

Subsequently, multilinear models of pleasantness are constructed based on the results of the analysis of correlations. As in Section~\ref{sec:results_perc}, all combinations of the proposed physical variables are considered, and a VIF check is performed on predictors to ensure that no multi-collinearity between predictors exists in a model. On the present corpus, the best model in terms of $R^2_{adj}$ is:
\begin{equation}
\hat P_{1, \varphi} = 16.71 - 0.18~L_{50} + 1.01~\hat T_B(\alpha_{opt}, \beta_{opt}) \label{eq:pphi}
\end{equation}
where $\varphi$ indicates a model from physical variables. Indicators related to traffic and voices activity are absent compared to the perceptual models in Section~\ref{sec:results_perc}. This is because the $L_{50}$ used as a predictor is correlated to both traffic parameters $L_{T, p}$ and $T_{T, p}$ that appeared in perceptual models $\hat P_{1, p}$ (eq.\ref{eq:pp1}) and $\hat P_{2, p}$ (eq.\ref{eq:pp2}) respectively, and no strong contribution of objective variables representative of voices to pleasantness is identified in this corpus.

@@

Pleasantness prediction for simulated scenes using the detection model is as effective as the best baseline model from Section~\ref{sec:results_phys}, and outperforms previous models relying on physical indicators directly computed on the audio signal. Though, perceptual baseline models presented in Section~\ref{sec:results_perc} still display better overall performance metrics for all considered sub-corpora.
